Bayesian exponential random graph modeling of KWL Burial network

### 2020-04-15 analysis/paper/paper.Rmd
add code of post-Euro to rmd file and knit it

```{r tidy-data}
# combine some periods
burial_three_period_tidy <-
  burial %>%
  rename(burial_label = ID) %>%
  mutate(Phase = ifelse(Phase == 'euro', 'post', Phase)) %>%
  filter(!is.na(Phase)) %>%
  mutate(Gold_leaf = ifelse(Gold_leaf == "shatter", "1", Gold_leaf),
         Stoneware = ifelse(Stoneware == "base", "1", Stoneware),
         Stamped_ceramic = ifelse(Stamped_ceramic == "cluster", "1", Stamped_ceramic)) %>%
  mutate_at(21:ncol(.), as.numeric) %>%
  janitor::remove_empty(which = "cols") %>%
  mutate(total = rowSums(.[c(21:48, 50, 55, 56)], na.rm = TRUE)) %>% #prestige goods
  mutate(Porcelain = rowSums(.[c(40:48, 55, 56)], na.rm = TRUE)) %>% #B&W, porcelain, Anping, kendi
  mutate(Porcelain = ifelse(Porcelain == 0, NA, Porcelain)) %>%
  mutate(quantity = case_when(
    total == 0 ~ "none",
    total > 0 & total <= 7 ~ "low",
    total > 7 & total < 100 ~ "medium",
    total >= 100 ~ "high",
    TRUE ~ "other")) %>% # the classification is based on the result of histogram
  mutate(Age_scale = case_when(
    `Age` %in% c("1","2") ~ "0-12",
    `Age` == "3" ~ "12~20",
    `Age` %in% c("4","5","6","7","8") ~ "+20",
    TRUE ~ "NA")) %>%
  mutate(Gold_bead_low = ifelse(Golden_bead == 1, 1, NA),
         Gold_bead_med = ifelse(Golden_bead > 1 & Golden_bead <10, 1, NA),
         Gold_bead_high = ifelse(Golden_bead > 10, 1, NA),
         Agate_bead_low = ifelse(Agate_bead == 1, 1, NA),
         Agate_bead_med = ifelse(Agate_bead > 1 & Agate_bead <10, 1, NA),
         Agate_bead_high = ifelse(Agate_bead > 10, 1, NA),
         `Indo-Pacific_bead_low` = ifelse(`Indo-Pacific_bead` < 100, 1, NA),
         `Indo-Pacific_bead_med` = ifelse(`Indo-Pacific_bead` > 100 & `Indo-Pacific_bead` < 900, 1, NA),
         `Indo-Pacific_bead_high` = ifelse(`Indo-Pacific_bead` > 900, 1, NA)) %>% 
  # the low-med-high index is determined by interpretation for histograms for those burial good types
  select(burial_label,
         Phase,
         Age_scale,
         Gold_bead_low,
         Gold_bead_med,
         Gold_bead_high,
         Agate_bead_low,
         Agate_bead_med,
         Agate_bead_high,
         #Agate_bead, #female burials
         #Golden_bead,
         Porcelain,
         Gold_leaf, #prestige good
         fish_shape_knit, #prestige good
         #Bell, #children's burials
         quantity)
         #total) # select specific variable to drop columns (uninformative variables)

@@

```{r network-object-pre}
# filter pre burials
burial_pre <-
  burial_three_period_tidy %>%
  filter(Phase == "pre") %>%
  janitor::remove_empty(which = "cols")

@@

# create list for each burial that contains the burial good types and their counts
edge_list_pre <-
  burial_three_period_tidy %>%
  select(burial_label, 4:12) %>% # need to change for each exploration
  pivot_longer(-burial_label, names_to = "goods", values_to = "count") %>%
  #mutate(burial_connection = rep(unique(burial_label), length.out = length(burial_label)))
  group_by(burial_label) %>%
  nest()

@@

@@

```{r network-ggraph}
# network analysis using ggraph pkg
library(tidygraph)
library(ggraph)

relation_tidy_pre <- tbl_graph(nodes = nodes_pre,
                               edges = edges_for_network_pre,
                               directed = FALSE)

@@

# plot
ggraph(relation_tidy_pre, layout = "graphopt") +
  geom_node_point() +
  geom_edge_link(aes(width = common_counts), alpha = 0.8) +
  scale_edge_width(range = c(0.2, 2)) +
  geom_node_text(aes(label = burial_label), repel = TRUE) +
  labs(edge_width = "common item") +
  theme_graph()
```

```{r network-ggraph}
# network analysis using ggraph pkg
library(tidygraph)
library(ggraph)

relation_tidy_pre <- tbl_graph(nodes = nodes_pre,
                               edges = edges_for_network_pre,
                               directed = FALSE)

# plot
ggraph(relation_tidy_pre, layout = "graphopt") +
  geom_node_point() +
  geom_edge_link(aes(width = common_counts), alpha = 0.8) +
  scale_edge_width(range = c(0.2, 2)) +
  geom_node_text(aes(label = burial_label), repel = TRUE) +
  labs(edge_width = "common item") +
  theme_graph()
```

```{r network-network-pkg}
# network analysis using network pkg
library(network)

@@

# plot
plot(burial_network_pre, vertex.cex = 1) # last two nodes are missing?? only 27 nodes
```

@@

# Add attributes to the network object burial_network_pre
set.vertex.attribute(burial_network_pre, "quantity", burial_pre$quantity)
set.vertex.attribute(burial_network_pre, "age", burial_pre$Age_scale)

@@

ID <- get.vertex.attribute(burial_network_pre, "burial_label") # not sure how to get id on the network plot

plot(burial_network_pre,
     vertex.col = "quantity", # age
     vertex.cex = 1.5)

@@

#? can't get the items in legend in order, tried as.factor and set their level but does not work
#? can't match color with categories, using an odd method at the moment
```

@@

# model 3 with node attributes
model.3 <- burial_network_pre ~ edges +  # the overall density of the network
  nodematch('quantity') + # quantity-based homophily, categorical nodal attribute, the similarity of connected nodes
  nodematch('age') +
  gwesp(0.2, fixed = TRUE) +    # transitivity
  gwdegree(0.8, fixed = TRUE)   # popularity
summary(model.3)
```

@@

```{r Bayesian-ERGMs}
# Specify a prior distribution: normal distribution (low density and high transitivity)
# need to observe the graph and select suitable prior 
prior.mean <- c(1, 0, 0, 3, 0) # prior mean corresponds to mean for each parameter
# follow Alberto Caimo et al. (2015) hospital example
prior.sigma <- diag(3, 5, 5) # covariance matrix structure
# where the dimension d corresponds to the number of parameters, 𝜇 is mean vector and Σprior is a d × d covariance matrix.

# Estimated posterior means, medians and 95% credible intervals for Models.3
# bergmM: Bayesian exponential random graphs models under missing data using the approximate exchange algorithm. Use this because of missing data for age variable
parpost <- bergmM(model.3,
                  prior.mean  = prior.mean,
                  prior.sigma = prior.sigma,
                  burn.in     = 200, # burn-in iterations for every chain of the population, drops the first 200
                  main.iters  = 2000, # iterations for every chain of the population
                  aux.iters   = 10000, # MCMC steps used for network simulation
                  nchains     = 8, # number of chains of the population MCMC
                  gamma       = 0.7) # scalar; parallel adaptive direction sampling move factor, acceptance rate

@@

```{r Bayesian-gof}
# Model assessment, Bayesian goodness of fit diagnostics:
bgof(parpost,
     aux.iters = 10000,
     n.deg     = 14,
     n.dist    = 15,
     n.esp     = 9)

Figure shows the Bayesian goodness of fit diagnostics plots where the red lines represent the distributions of the observed data and the box plots represent the GOF distributions of 10000 network graphs simulated from ERGMs based on the estimated posterior distribution. The grey lines mark the 95% intervals.

```{r Bayesian-gof}
# Model assessment, Bayesian goodness of fit diagnostics:
bgof(parpost,
     aux.iters = 10000,
     n.deg     = 14,
     n.dist    = 15,
     n.esp     = 9)

>>>>

```{r tidy-data}
# combine some periods
burial_three_period_tidy <-
  burial %>%
  rename(burial_label = ID) %>%
  mutate(Phase = ifelse(Phase == 'euro', 'post', Phase)) %>%
  filter(!is.na(Phase)) %>%
  mutate(Gold_leaf = ifelse(Gold_leaf == "shatter", "1", Gold_leaf),
         Stoneware = ifelse(Stoneware == "base", "1", Stoneware),
         Stamped_ceramic = ifelse(Stamped_ceramic == "cluster", "1", Stamped_ceramic)) %>%
  mutate_at(21:ncol(.), as.numeric) %>%
  janitor::remove_empty(which = "cols") %>%
  mutate(total = rowSums(.[c(21:48, 50, 55, 56)], na.rm = TRUE)) %>% #prestige goods
  mutate(Porcelain = rowSums(.[c(40:48, 55, 56)], na.rm = TRUE)) %>% #B&W, porcelain, Anping, kendi
  mutate(Porcelain = ifelse(Porcelain == 0, NA, Porcelain)) %>%
  mutate(quantity = case_when(
    total == 0 ~ "none",
    total > 0 & total <= 7 ~ "low",
    total > 7 & total < 100 ~ "medium",
    total >= 100 ~ "high",
    TRUE ~ "other")) %>% # the classification is based on the result of histogram
  mutate(Age_scale = case_when(
    `Age` %in% c("1","2") ~ "0-12",
    `Age` == "3" ~ "12~20",
    `Age` %in% c("4","5","6","7","8") ~ "+20",
    TRUE ~ "NA")) %>%
    mutate(gender = case_when(
    `Gender` %in% c("1","2") ~ "male",
    `Gender` %in% c("3","4") ~ "female",
    TRUE ~ "NA")) %>%
  mutate(Gold_bead_low = ifelse(Golden_bead == 1, 1, NA),
         Gold_bead_med = ifelse(Golden_bead > 1 & Golden_bead <10, 1, NA),
         Gold_bead_high = ifelse(Golden_bead > 10, 1, NA),
         Agate_bead_low = ifelse(Agate_bead == 1, 1, NA),
         Agate_bead_med = ifelse(Agate_bead > 1 & Agate_bead <10, 1, NA),
         Agate_bead_high = ifelse(Agate_bead > 10, 1, NA),
         `Indo-Pacific_bead_low` = ifelse(`Indo-Pacific_bead` < 100, 1, NA),
         `Indo-Pacific_bead_med` = ifelse(`Indo-Pacific_bead` > 100 & `Indo-Pacific_bead` < 900, 1, NA),
         `Indo-Pacific_bead_high` = ifelse(`Indo-Pacific_bead` > 900, 1, NA)) %>% 
  # the low-med-high index is determined by interpretation for histograms for those burial good types
  select(burial_label,
         Phase,
         Age_scale,
         gender,
         Gold_bead_low,
         Gold_bead_med,
         Gold_bead_high,
         Agate_bead_low,
         Agate_bead_med,
         Agate_bead_high,
         #Agate_bead, #female burials
         #Golden_bead,
         Porcelain,
         Gold_leaf, #prestige good
         fish_shape_knit, #prestige good
         #Bell, #children's burials
         quantity)
         #total) # select specific variable to drop columns (uninformative variables)

@@

```{r prep-network-object-pre}
# filter pre burials
burial_pre <-
  burial_three_period_tidy %>%
  filter(Phase == "pre") %>%
  janitor::remove_empty(which = "cols")

@@

# create list for each burial that contains the burial good types and their counts
edge_list_pre <-
  burial_three_period_tidy %>%
  select(burial_label, 5:13) %>% # need to change for each exploration
  pivot_longer(-burial_label, names_to = "goods", values_to = "count") %>%
  #mutate(burial_connection = rep(unique(burial_label), length.out = length(burial_label)))
  group_by(burial_label) %>%
  nest()

@@

```{r prep-network-object-post}
# filter post burials
burial_post <-
  burial_three_period_tidy %>%
  filter(Phase == "post") %>%
  janitor::remove_empty(which = "cols")

# create node list
nodes_post <-
  burial_three_period_tidy %>%
  filter(Phase == "post") %>%
  select(burial_label) %>%
  rowid_to_column("id")

# pair wise combinations for burials as index for later map function
library(gtools)
burial_comb_post = combinations(length(nodes_post$burial_label),
                               2, nodes_post$burial_label,
                               repeats = TRUE)
colnames(burial_comb_post) = c("burial_1", "burial_2")
burial_comb_post = as_tibble(burial_comb_post)

# create list for each burial that contains the burial good types and their counts
edge_list_post <-
  burial_three_period_tidy %>%
  select(burial_label, 5:13) %>% # need to change for each exploration
  pivot_longer(-burial_label, names_to = "goods", values_to = "count") %>%
  group_by(burial_label) %>%
  nest()

# pair-wise list
common_counts_lst_post <-
  map2(burial_comb_post$burial_1,
       burial_comb_post$burial_2,
       ~bind_cols(
         edge_list_post %>%
           filter(burial_label == .x) %>%
           unnest(data),
         edge_list_post %>%
           filter(burial_label == .y) %>%
           unnest(data)) %>%
         rowwise() %>%
         mutate(common_counts = sum(count, count1)))

# count of ornament types in common between each pair of burials
common_counts_vct_post <- map_int(common_counts_lst_post, ~sum(!is.na(.x$common_counts)))

burial_comb_with_common_counts_post <-
  burial_comb_post %>%
  mutate(common_counts = common_counts_vct_post) %>%
  mutate(common_counts = ifelse(burial_1 == burial_2, 0, common_counts)) # use 0 for self loop

# change label to ids for node linking
edges_post <-
  burial_comb_with_common_counts_post %>%
  left_join(nodes_post, by = c("burial_1" = "burial_label")) %>%
  rename(from = id) %>%
  left_join(nodes_post, by = c("burial_2" = "burial_label")) %>%
  rename(to = id)

edges_for_network_post <-
  select(edges_post, from, to, common_counts) %>%
  filter (!common_counts == 0) # remove rows with no goods in common
```

@@

```{r network-ggraph}
# network analysis using ggraph pkg
library(tidygraph)
library(ggraph)
# pre-Euro
relation_tidy_pre <- tbl_graph(nodes = nodes_pre,
                               edges = edges_for_network_pre,
                               directed = FALSE)

@@

# plot
ggraph(relation_tidy_pre, layout = "graphopt") +
  geom_node_point() +
  geom_edge_link(aes(width = common_counts), alpha = 0.8) +
  scale_edge_width(range = c(0.2, 2)) +
  geom_node_text(aes(label = burial_label), repel = TRUE) +
  labs(edge_width = "common item") +
  theme_graph()

# post-Euro
relation_tidy_post <- tbl_graph(nodes = nodes_post,
                               edges = edges_for_network_post,
                               directed = FALSE)

relation_tidy_post %>%
  activate(edges) %>%
  arrange(desc(common_counts))

# plot
ggraph(relation_tidy_post, layout = "graphopt") +
  geom_node_point() +
  geom_edge_link(aes(width = common_counts), alpha = 0.8) +
  scale_edge_width(range = c(0.2, 2)) +
  geom_node_text(aes(label = burial_label), repel = TRUE) +
  labs(edge_width = "common item") +
  theme_graph()
```

```{r network-object}
library(network)

@@

burial_network_post <-
  network(edges_for_network_post,
          vertex.attr = nodes_post,
          directed = FALSE,
          ignore.eval = FALSE, 
          loops = FALSE, 
          matrix.type = "edgelist") 
```

@@

# Add attributes to the network object burial_network_pre
set.vertex.attribute(burial_network_pre, "quantity", burial_pre$quantity)
set.vertex.attribute(burial_network_pre, "age", burial_pre$Age_scale)
set.vertex.attribute(burial_network_pre, "gender", burial_pre$gender)

@@

plot(burial_network_pre,
     displaylabels = TRUE,
     vertex.col = "quantity",
     vertex.cex = degree(burial_network_pre, cmode = 'indegree')/5, #size nodes
     pad = 1)

@@

#? can't get the items in legend in order, tried as.factor and set their level but does not work
#? can't match color with categories, using an odd method at the moment

# Add attributes to the network object burial_network_post
set.vertex.attribute(burial_network_post, "quantity", burial_post$quantity)
set.vertex.attribute(burial_network_post, "age", burial_post$Age_scale)
set.vertex.attribute(burial_network_post, "gender", burial_post$gender)

# plot
set.seed(30)
quantity <- get.vertex.attribute(burial_network_post, "quantity")
age <- get.vertex.attribute(burial_network_post, "age")
ID <- get.vertex.attribute(burial_network_post, "burial_label") # not sure how to get id on the network plot

plot(burial_network_post,
     displaylabels = TRUE,
     vertex.col = "quantity",
     vertex.cex = degree(burial_network_post, cmode = 'indegree') / 13, #size nodes to their in-degree
     #vertex.sides = ifelse(burial_network_pre %v% "", 4, 50),
     pad = 1)

legend("topleft",
       col = c(4, 3, 2, 1), # need to adjust each time
       pch    = 20,
       legend = unique(quantity),
       title  = 'Burial good counts')
```

@@

# model 3 with node attributes
model.3 <- burial_network_pre ~ edges +  # the overall density of the network
  nodematch('quantity') + # quantity-based homophily
  nodematch('age') +
  nodematch('gender') +
  gwesp(0.75, fixed = TRUE) +    # transitivity
  gwnsp(0.75, fixed = TRUE) +
  gwdegree(0.2, fixed = TRUE) # popularity
summary(model.3)
```

@@

```{r Bayesian-ERGMs}
# Specify a prior distribution: normal distribution (low density and high transitivity)
# need to observe the graph and select suitable prior 
prior.mean <- c(2, 0, 0, 0, 5, 0, 0) 
# follow Alberto Caimo et al. (2015) hospital example
prior.sigma <- diag(3, 7) # covariance matrix structure
# where the dimension d corresponds to the number of parameters, 𝜇 is mean vector and Σprior is a d × d covariance matrix.

# Estimated posterior means, medians and 95% credible intervals for Models.3
# bergmM: Bayesian exponential random graphs models under missing data using the approximate exchange algorithm. Use this because of missing data for age variable
parpost <- bergmM(model.3,
                  prior.mean  = prior.mean,
                  prior.sigma = prior.sigma,
                  burn.in     = 200, # burn-in iterations for every chain of the population, drops the first 200
                  main.iters  = 2000, # iterations for every chain of the population
                  aux.iters   = 10000, # MCMC steps used for network simulation
                  nchains     = 8, # number of chains of the population MCMC
                  gamma       = 0.2) # scalar; parallel adaptive direction sampling move factor, acceptance rate

@@

```{r Bayesian-gof}
# Model assessment, Bayesian goodness of fit diagnostics:
bgof(parpost,
     aux.iters = 10000,
     n.deg     = 15,
     n.dist    = 15,
     n.esp     = 15)
